The Impact of Alternation

### 2017-05-14 draft.tex
appendix

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{blast} and the \impact~method
\cite{mcmillan06}. These algorithms are proven to terminate and return
a word from the language, if the language is not empty, but
termination is not guaranteed when the language is empty. Finally, we
prove termination in the empty case, when the transition
formulae\begin{inparaenum}[(i)]
\item contain only constants of the data sort and boolean functions of
  any arity, or they
%
\item constrain only the current but not the past values on the input,
  as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
\end{inparaenum}

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
\end{array}\]
It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
every formula $\phi \in \Form^+(Q,\vec{x})$. If $\phi$ does not
contain boolean variables, then $\overline{\phi}$ and $\neg\phi$ are
equivalent, and in general:

@@

\begin{lemma}\label{lemma:closure}
  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
  L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
  L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) = \Sigma[\vec{x}]^*
  \setminus L(\A_1)$. \qed}

@@

\begin{lemma}\label{lemma:safety-invariant}
  For any automaton $\A$, we have $L(\A) = \emptyset$ if and only if
  $\A$ has a safety invariant.
\end{lemma}
\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
  u \in \Sigma^* ~.~ \beta\cup\nu \models
  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
  straightforward. \qed}

@@

Formally, we fix a set of formulae $\preds \subseteq
\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
formulae as \emph{predicates}. Given a formula $\phi$, we denote by
$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
$\preds$. The abstract versions of the post-image and acceptance
condition are defined as follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

\begin{proposition}\label{prop:positive-interpol}
  For any $u = a_1 \ldots a_n \in \Sigma^*$, if $\pi(u)$ is
  unsatisfiable then one can build an interpolant
  $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\pi(u)$, such that $I_i \in
  \Form^+(Q,\vec{x})$, for all $i \in [0,n]$.
\end{proposition}
\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
  and $i \in [0,n]$ be the first index for which $I_i \not\in
  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
  an interpolant, because: \begin{compactitem}
  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
    %
  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
  \end{compactitem}
  We proceed with these replacements until there are no more formulae
  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}

@@

\begin{lemma}\label{lemma:progress}
  For any non-empty sequence $u = a_1\ldots a_n \in \Sigma^*$
  such that $\Accept{\A}(u)$ is unsatisfiable, the following
  hold: \begin{compactenum}
  \item\label{it1:progress} $\Theta(u)$ is unsatisfiable, and
    %
  \item\label{it2:progress} if $\tuple{\top,I_0,\dots,I_{n},\bot}$ is
    an interpolant for $\Theta(u)$ such that $\set{I_i \mid i
      \in[0,n]} \subseteq \preds$ then $\AbsAccept{\A}(u)$ is
    unsatisfiable.
  \end{compactenum}
\end{lemma}
\proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
  Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1}
  \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)$ in
  the following.

  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
  interpolant for $\pi(u)$, the following entailments
  hold: \begin{compactitem}
  \item $\theta_0 \models I_0[Q_0/Q]$,
    %
  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
    %
  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
  \end{compactitem}
  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
  induction on $n \geq 0$. This is sufficient to conclude because
  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
  \models I_0$. For the induction step $n>0$, we compute:
  \[\begin{array}{l}
  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
  \end{array}\]}
We have now all the ingredients to describe the first emptiness
semi-algorithm for alternating data
automata. Algorithm\footnote{Though termination is not guaranteed, we
  call it algorithm for conciseness.} \ref{alg:predabs} builds an
\emph{abstract reachability tree} (ART) whose nodes are labeled with
formulae over-approximating the concrete sets of configurations, which
uses a covering relation between nodes in order to ensure that the set
of formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample is eliminated by computing an interpolant and
adding its formulae to the set of predicates.

@@

\begin{theorem}\label{thm:predabs}
  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
  \ref{alg:predabs} terminates and returns a word $w \in L(\A)$. If
  Algorithm \ref{alg:predabs} terminates reporting $\true$, then
  $L(\A) = \emptyset$.
\end{theorem}
\proof{We prove the following invariant: each time Algorithm
  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
  in $\worklist$ contains all the frontier nodes in the ART
  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
  some node in $N$, namely that:
  \begin{equation}\label{inv:predabs}
  W = \set{n \not\in N \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
    \not\in E ~\wedge~ (n,m) \not\in \lhd}
  \end{equation}
  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
  = \lhd = \emptyset$. If the invariant holds previously, at line
  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
  executed, because, when the subtree rooted at the pivot $p$ is
  removed, $p$ becomes a member of the set of uncovered frontier
  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
  considers that the invariant holds at line \ref{ln:while} and the
  control follows the else branch at line \ref{ln:else}. In this case,
  the newly created frontier node $s$ is added to $W$ only if it is
  not covered by an existing node in $N$ (line \ref{ln:covered}).

@@

\begin{lemma}\label{lemma:well-labeled}
  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
  \Lambda(m)$, for all $(n,a,m) \in E$.
\end{lemma}
\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
  the entailment holds trivially. Otherwise, let $\Omega$ be the set
  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
  be a spurious counterexample. Let
  $\langle\top,I^\omega_0,\ldots,I^\omega_k\equiv\bot\rangle$ be an
  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
  \ref{prop:positive-interpol}, it is possible to build such an
  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
  Proposition \ref{prop:asterisk}, we obtain
  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}

@@

\begin{theorem}\label{thm:impact}
  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
  \ref{alg:impact} terminates and returns a word $w \in L(\A)$. If
  Algorithm \ref{alg:impact} terminates reporting $\true$, then
  $L(\A) = \emptyset$.
\end{theorem}
\proof{ We prove first that, each time Algorithm \ref{alg:impact}
  reaches the line \ref{ln:impact-while}, we have:
  \begin{equation}\label{inv:impact}
  W = \set{n \not\in N \mid n \text{ uncovered, } \exists a \in
    \Sigma~ \forall s \in N ~.~ (n,a,s) \not\in E}
  \end{equation}
  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus
  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
  holds at when reaching line \ref{ln:impact-while} and some node $n$
  was removed from $W$ and inserted into $N$. We distinguish two
  cases, either: \begin{itemize}
  \item $n$ is covered, in which case $W$ becomes $W \setminus
    \set{n}$, or
  %
  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
      a \in \Sigma}$ is the set of fresh successors of $n$. But then
    no node $s \in S$ is covered and has successors in $E$.
  \end{itemize}
  Then the condition (\ref{inv:impact}) holds next time line
  \ref{ln:impact-while} is reached, thus it is invariant.

>>>>

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{blast} and the \impact~method
\cite{mcmillan06}.  These algorithms are proven to terminate and
return a word from the language, if the language is not empty, but
termination is not guaranteed when the language is empty.

%% Finally, we prove termination in the empty case, when the transition
%% formulae\begin{inparaenum}[(i)]
%% \item contain only constants of the data sort and boolean functions of
%%   any arity, or they
%% %
%% \item constrain only the current but not the past values on the input,
%%   as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
%% \end{inparaenum}

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
\end{array}\]
It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
every formula $\phi \in \Form^+(Q,\vec{x})$.

@@

\begin{lemma}\label{lemma:closure}
  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
  L(\A_1)$.
\end{lemma}

@@

\begin{lemma}\label{lemma:safety-invariant}
  For any automaton $\A$, we have $L(\A) = \emptyset$ if and only if
  $\A$ has a safety invariant.
\end{lemma}

@@

Formally, we fix a set of formulae $\preds \subseteq
\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
formulae as \emph{predicates}. Given a formula $\phi$, we denote by
$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
$\preds$. The abstract versions of the post-image and acceptance
condition are defined as follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]

\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}

@@

\begin{proposition}\label{prop:positive-interpol}
  For any $u = a_1 \ldots a_n \in \Sigma^*$, if $\pi(u)$ is
  unsatisfiable then one can build an interpolant
  $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\pi(u)$, such that $I_i \in
  \Form^+(Q,\vec{x})$, for all $i \in [0,n]$.
\end{proposition}

@@

\begin{lemma}\label{lemma:progress}
  For any non-empty sequence $u = a_1\ldots a_n \in \Sigma^*$
  such that $\Accept{\A}(u)$ is unsatisfiable, the following
  hold: \begin{compactenum}
  \item\label{it1:progress} $\Theta(u)$ is unsatisfiable, and
    %
  \item\label{it2:progress} if $\tuple{\top,I_0,\dots,I_{n},\bot}$ is
    an interpolant for $\Theta(u)$ such that $\set{I_i \mid i
      \in[0,n]} \subseteq \preds$ then $\AbsAccept{\A}(u)$ is
    unsatisfiable.
  \end{compactenum}
\end{lemma}

We have now all the ingredients to describe the first emptiness
semi-algorithm for alternating data
automata. Algorithm\footnote{Though termination is not guaranteed, we
  call it algorithm for conciseness.} \ref{alg:predabs} builds an
\emph{abstract reachability tree} (ART) whose nodes are labeled with
formulae over-approximating the concrete sets of configurations, which
uses a covering relation between nodes in order to ensure that the set
of formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample is eliminated by computing an interpolant and
adding its formulae to the set of predicates.

@@

\begin{theorem}\label{thm:predabs}
  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
  \ref{alg:predabs} terminates and returns a word $w \in L(\A)$. If
  Algorithm \ref{alg:predabs} terminates reporting $\true$, then
  $L(\A) = \emptyset$.
\end{theorem}

@@

\begin{lemma}\label{lemma:well-labeled}
  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
  \Lambda(m)$, for all $(n,a,m) \in E$.
\end{lemma}

@@

\begin{theorem}\label{thm:impact}
  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
  \ref{alg:impact} terminates and returns a word $w \in L(\A)$. If
  Algorithm \ref{alg:impact} terminates reporting $\true$, then
  $L(\A) = \emptyset$.
\end{theorem}

\appendix

\section{Proofs}

\subsection{Proof of Lemma \ref{lemma:closure}}

\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
  L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) = \Sigma[\vec{x}]^*
  \setminus L(\A_1)$. \qed}

\subsection{Proof of Lemma \ref{lemma:safety-invariant}}

\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
  u \in \Sigma^* ~.~ \beta\cup\nu \models
  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
  straightforward. \qed}

\subsection{Proof of Lemma \ref{lemma:abstract-invariant}}

\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

\subsection{Proof of Proposition \ref{prop:positive-interpol}}

\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
  and $i \in [0,n]$ be the first index for which $I_i \not\in
  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
  an interpolant, because: \begin{compactitem}
  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
    %
  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
  \end{compactitem}
  We proceed with these replacements until there are no more formulae
  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}

\subsection{Proof of Lemma \ref{lemma:progress}}
